\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-05-10 paper.tex
Remove dependent of beta in survival formulas

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{ x^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, \beta_k^l, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
the history of the true unobserved longitudinal process up to time $t$. We choose to incorporate $x$, also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of time-independent features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independent regularizations are used during inference, see~\eqref{eq:pen-log-lik}. Corresponding fixed effects are denoted $\gamma_k^0 \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$. 
This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
\[({\gamma_k^0}^\top, \gamma_k^\top)^\top= ({\gamma_k^0}^\top, \gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]

@@

The association structure between the longitudinal and the time-to-event submodels is key to the joint modeling framework. In spite of that, rationale for selecting shared associations has received little attention. We then propose to include some of the most common parameterization with no \textit{a priori}, set out in Table~\ref{table:shared_associations}, and let the model select the relevant ones through the regularization strategy described in Section~\ref{sec:penalized-obj}.
\begin{table}[htb]
\centering
\begin{tabular}{ccccc}
\toprule
Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\frac{\partial \varphi_a(t, \beta_k^l, b^l)}{\partial \beta_k^l}$ & $\imath_a$ & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & $u^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$ & $\mathbf{0}_{q_l}$ & $r_l$ & \citet{hatfield2011joint} \\ 
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & $\dfrac{\dd}{\dd t} u^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & $\int_0^t u^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Description of the shared associations included in the group-specific risk of event submodel~\eqref{eq:intensity-model}. The gradient with respect to $\beta_k^l$ is also given, which will be useful for inference in Section~\ref{sec:QNMCEM}.}
\label{table:shared_associations}
\end{table}

@@

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + x_i^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, \beta_k^l, b_i^l) \Big) \\
  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ x_i^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l) \Big\} \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.

@@

\paragraph*{Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \big(\norm{\xi_k}_{\text{en}, \eta} + \norm{\gamma_k^0}_{\text{en}, \eta} \big) + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
Let us introduce the following useful notations
\begin{equation}
  \label{eq:g-tilde-1-2}
  \left\{
    \begin{array}{ll}
        \tilde g_{i, s, \gamma_k^0, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_k^0 + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\}, \\
        \tilde g_{s, \gamma_k, \beta_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l).
    \end{array}
  \right. 
\end{equation}
We then present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\hat \pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}. Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:QNMCEM}) with every increase of $N$.

@@

Concerning the update of ${\gamma_k^0}^{(w)}$ and $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k^0, \gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k^0, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] 
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k^0$ and $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k^0, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}. The update for ${\gamma_k^0}^{(w)}$ therefore requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k^0}
  {\gamma_k^0}^{(w+1)} \in \argmin_{\gamma_k^0 \in \R^p} Q^{(w)}_{n,k}(\gamma_k^0, {\gamma_k}^{(w)}) + \zeta_{1,k} \norm{\gamma_k^0}_{\text{en}, \eta}.
\end{equation}
We then rewrite problem~\eqref{eq:minimization-gamma_k^0} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-gamma_k^0}
  \begin{split}
    \text{minimize}& \quad \quad Q_{n, k}^{(w)}({\gamma_k^0}^+ - {\gamma_k^0}^-, {\gamma_k}^{(w)}) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p ({\gamma_{k,j}^0}^+ + {\gamma_{k,j}^0}^-) + \dfrac \eta 2 \norm{{\gamma_k^0}^+ - {\gamma_k^0}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad {\gamma_{k,j}^0}^+ \geq 0 \text{ and } {\gamma_{k,j}^0}^- \geq 0 \text{ for } j = 1, \ldots, p.
  \end{split} 
\end{equation}
The gradient is here given by
\begin{align}
  \label{eq:grad-Q-1}
  \frac{\partial Q_{n,k}^{(w)}(\gamma_k^0, {\gamma_k}^{(w)})}{\partial \gamma_k^0} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \nonumber \\
  &\times \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, {\gamma_k}^{(w)}, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i.
\end{align}
Then, the update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}({\gamma_k^0}^{(w+1)}, \gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}
We treat problem~\eqref{eq:minimization-gamma_k} in the same way as~\eqref{eq:minimization-beta_k}, that is
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}({\gamma_k^0}^{(w+1)}, \cdot), \tilde \eta, \zeta_{3,k}\big)
\end{equation}
with the gradient being known since
\begin{align}
  \label{eq:grad-Q-2}
  &\frac{\partial Q_{n,k}^{(w)}({\gamma_k^0}^{(w+1)}, \gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, {\beta_k^l}^{(w+1)}, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{i, t_j, {\gamma_k^0}^{(w+1)}, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

For the update of $\lambda_0^{(w)}$, we treat the jump size at the observed event times as parameters to be estimated~\citep{klein1992semiparametric}. The closed-form update is then given by
\begin{equation}
  \label{eq:update_lambda_0}
  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[ \tilde g_{i, t, \gamma_k^{(w+1)}, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
\end{equation}
which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.

>>>>

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{ x^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
the history of the true unobserved longitudinal process up to time $t$. We choose to incorporate $x$, also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of time-independent features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independent regularizations are used during inference, see~\eqref{eq:pen-log-lik}. Corresponding fixed effects are denoted $\gamma_k^0 \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$. 
This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
\[({\gamma_k^0}^\top, \gamma_k^\top)^\top= ({\gamma_k^0}^\top, \gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]

@@

The association structure between the longitudinal and the time-to-event submodels is key to the joint modeling framework. In spite of that, rationale for selecting shared associations has received little attention. We then propose to include some of the most common parameterization with no \textit{a priori}, set out in Table~\ref{table:shared_associations}, and let the model select the relevant ones through the regularization strategy described in Section~\ref{sec:penalized-obj}.
\begin{table}[htb]
\centering
\begin{tabular}{ccccc}
\toprule
Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\imath_a$ & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Description of the shared associations included in the group-specific risk of event submodel~\eqref{eq:intensity-model}.}
\label{table:shared_associations}
\end{table}

@@

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + x_i^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, b_i^l) \Big) \\
  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ x_i^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l) \Big\} \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.

@@

\paragraph*{Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \big(\norm{\xi_k}_{\text{en}, \eta} + \norm{\gamma_k^0}_{\text{en}, \eta} \big) + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
Let us introduce the following useful notations
\begin{equation}
  \label{eq:g-tilde-1-2}
  \left\{
    \begin{array}{ll}
        \tilde g_{i, s, \gamma_k^0, \gamma_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_k^0 + \tilde g_{s, \gamma_k}^2(b_i) \big\}, \\
        \tilde g_{s, \gamma_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l).
    \end{array}
  \right. 
\end{equation}
We then present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\hat \pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}. Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:QNMCEM}) with every increase of $N$.

@@

Concerning the update of ${\gamma_k^0}^{(w)}$ and $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k^0, \gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k^0, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k^0$ and $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k^0, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}. The update for ${\gamma_k^0}^{(w)}$ therefore requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k^0}
  {\gamma_k^0}^{(w+1)} \in \argmin_{\gamma_k^0 \in \R^p} Q^{(w)}_{n,k}(\gamma_k^0, {\gamma_k}^{(w)}) + \zeta_{1,k} \norm{\gamma_k^0}_{\text{en}, \eta}.
\end{equation}
We then rewrite problem~\eqref{eq:minimization-gamma_k^0} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-gamma_k^0}
  \begin{split}
    \text{minimize}& \quad \quad Q_{n, k}^{(w)}({\gamma_k^0}^+ - {\gamma_k^0}^-, {\gamma_k}^{(w)}) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p ({\gamma_{k,j}^0}^+ + {\gamma_{k,j}^0}^-) + \dfrac \eta 2 \norm{{\gamma_k^0}^+ - {\gamma_k^0}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad {\gamma_{k,j}^0}^+ \geq 0 \text{ and } {\gamma_{k,j}^0}^- \geq 0 \text{ for } j = 1, \ldots, p.
  \end{split} 
\end{equation}
The gradient is here given by
\begin{align}
  \label{eq:grad-Q-1}
  \frac{\partial Q_{n,k}^{(w)}(\gamma_k^0, {\gamma_k}^{(w)})}{\partial \gamma_k^0} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \nonumber \\
  &\times \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, {\gamma_k}^{(w)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i.
\end{align}
Then, the update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}({\gamma_k^0}^{(w+1)}, \gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}
We treat problem~\eqref{eq:minimization-gamma_k} in the same way as~\eqref{eq:minimization-beta_k}, that is
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}({\gamma_k^0}^{(w+1)}, \cdot), \tilde \eta, \zeta_{3,k}\big)
\end{equation}
with the gradient being known since
\begin{align}
  \label{eq:grad-Q-2}
  &\frac{\partial Q_{n,k}^{(w)}({\gamma_k^0}^{(w+1)}, \gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \tilde g_{i, t_j, {\gamma_k^0}^{(w+1)}, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

For the update of $\lambda_0^{(w)}$, we treat the jump size at the observed event times as parameters to be estimated~\citep{klein1992semiparametric}. The closed-form update is then given by
\begin{equation}
  \label{eq:update_lambda_0}
  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[ \tilde g_{i, t, \gamma_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
\end{equation}
which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.
